Type theories as quotient inductive-recursive types

### 2025-02-14 tex/main.tex
Minor revisions before adding definitions in the appendix

\subparagraph*{Plan of the paper}
In the rest of the paper, we introduce the metatheory in \cref{sec:meta}.
Then, we will develop type theories formulated as QIIRTs in \cref{sec:QIIRTs}, followed by their elimination principles and usage in \cref{sec:models}.
We outline the \Agda formal implementation in \cref{sec:formal-implementation}, and we conclude with related and future work in \cref{sec:conclusion}.

@@

We work in intensional Martin-L\"of type theory with uniqueness of identity proofs. In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality and $\eta$-rules for $\Pi$-types and $\Sigma$-types.
%
We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B\,x$.
%
We are further inspired by Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an `implicit' argument which should be clear from the context.
We write $\Set_i$ for the cumulative hierarchy of universes with the level index $i$ suppressed.

@@

An ordinary inductive type $A$, such as for example the type of natural numbers $A = \mathbb{N}$, allows functions out of the type $f : A \to X$ to be defined by recursion, by providing an elimination rule which says that such a function $f$ is defined by its values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of $A$. For example, a function $f: \mathbb{N} \to X$ is given by the values $f(0)$ and $f(\mathsf{suc}\,n)$, where the latter might refer to the already computed value $f(n)$.
%
The type is first defined inductively by listing its constructors, and then afterwards recursive functions can be defined out of it.

In contrast, an inductive-recursive type is defined inductively \emph{simultaneously} with a recursively defined function out of it. The prototypical example is Martin-L\"of's universe \`a la Tarski, where a type $U$ of codes for small types is defined simultaneously with a `decoding' function $T : U \to \Set$ which maps each code to the type it represents. Crucially, the code for dependent function types
\[
  \pi : (a : U) \to (T(a) \to U) \to U
\]
with $T(\pi(a, b)) = (x : T(a)) \to T(b(x))$ only makes sense because we already have access to the decoding function $T$.
%
Inductive-recursive types (IRT) were introduced by Dybjer~\cite{Dybjer2000}, and then extensively studied by Dybjer and Setzer~\cite{Dybjer1999,Dybjer2003}.
The general schema allows one to define a type $A$ inductively, together with a recursive function $F : A \to D$, for some fixed (possibly large) type $D$.

An IRT $(A, F)$ comes with an elimination rule that allows to define a dependent function  $f : (x : A) \to P(x)$ by giving values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of $A$, again reflecting that $A$ is inductively defined.
%
However, note that the recursive function $F : A \to D$ is not defined using this elimination rule, but part of the definition of the type itself.
%
In particular, the elimination rule can be used to define endofunctions on $A$, but the recursive function  $F : A \to D$ must always target a fixed pre-existing type $D$.
%
In this work, we will consider an extension \emph{generalised} IRTs, where also $F : A \to A$ is allowed.
%
Such IRTs were previously used by Danielsson~\cite{Danielsson2006} to represent type theory in type theory, and are discussed by Kaposi~\cite{Kaposi2023}.
%
Their general metatheory would be interesting future work.

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020} from homotopy type theory, where constructors can not only introduce new elements, but also new equalities between the introduced elements.
%
Since we assume the uniqueness of identity proofs, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type of \emph{join lists} $\JList{A}$ can be given by the QIT
with the following constructors:
\begin{alignat*}{3}
  \Jnil                    & : \implicit{A}      && \JList{A} \\
  \Jsing{\blank}           & : \implicit{A}      && A \to \JList{A} \\
  \Jconcat{\blank}{\blank} & : \implicit{A}      && \JList{A} \to \JList{A} \to \JList{A} \\
  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
  \mathsf{assoc}           & : \implicit{\ell_0, \ell_1, \ell_2}\; && \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
\end{alignat*}

A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a family of types $B : A \to \Set$, where both $A$ and $B$ are defined inductively --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively; indeed, inductive-inductive definitions only make sense when $D = \Set$.
The metatheory of QIITs has been extensively studied by Altenkirch, Kaposi and Kov\'acs~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.

We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: The type $\JList{A}$ is defined simultaneously with a recursion $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list. We keep the constructors $\Jnil$, $\Jsing{\blank}$ and $\Jconcat{\blank}{\blank}$, but replace $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
\[
  \mathsf{nf} : \implicit{A}\;(\ell, \ell' : \JList{A}) \to \flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell' \to \ell =^{\JList{A}} \ell'
\]
which states that join lists are equal as soon as their flattenings are equal. For $\flatten$ to be well defined, it needs to respect the $\mathsf{nf}$ constructor, but this is a tautology: if $\ell =^{\JList{A}} \ell'$ because $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$, then obviously $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$.

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
\begin{figure} 
\begin{minipage}[t]{.48\textwidth}
  \begin{alignat*}{5}
    \blank  & + \blank     && : \N \to \N \to \N \\
    0       &  + m         && = m  \\
  n         &  + 0         && = m  \\
  (\suc\;n) &  + m         && = \suc\;(n + m) \\
  n         &  + (\suc\;m) && = \suc\;(n + m)
\end{alignat*}
\end{minipage}
\begin{minipage}[t]{.5\textwidth}
\begin{alignat*}{5}
  \blank         & +'\blank     && : \N \to \N \to \N \\
  0              & +' m         && \defeq m  \\
  (\suc\;n)      & +' 0         && \defeq \suc\;n \\
  (\suc\;n)      & +' (\suc\;m) && \defeq \suc\;(n +' \suc\;m)
\end{alignat*}
\end{minipage}
\caption{Two seemingly different definitions of addition}
\label{fig:additions}
\end{figure}
Functions are typically defined informally using pattern matching for convenience and readability. Formally, however, they are either defined using an explicit pattern-matching construct or elaborated into case trees, which correspond to the \emph{eliminator}, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination and completeness checking.
This elaboration offers the conventional notation, while making computational behaviour opaque and favouring asymmetric definitions.
For example, two definitions of addition in \cref{fig:additions} are the same
after elaboration in \Agda: eliminating the first argument of $+$ distinguishes $0$ and $\suc\;n$ and refines the second and the third clauses into clauses of $+'$, making the fourth clause useless.
Each clause of $+'$ is strict (hence the symbol $\defeq$ is used) while clauses of $+$ are not necessarily strict by definition.
%This semantics for function definitions decides how definitions reduce and therefore how identities are proved.

Cockx~\cite{Cockx2014} explores an alternative semantics for pattern matching, where all clauses are strict, even in the presence of \emph{overlapping patterns}.
Thus, when a function $f$ is defined with overlapping patterns, the notation $f\;\overrightarrow{p_i} \reduce t_i$ is used to emphasize its reduction behaviour.
Under this semantics, each clause of $+$ is strictly equal to its right-hand side, allowing the symmetry addition to be proved commutative directly by induction on $n$ and $m$.

@@

We begin with the definition of the parallel substitution calculus~\cite{Martin-Lof1992} as a QIIT, emphasising how the use of transports in its very definition complicates formal reasoning.
We then introduce its QIIRT counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning, and develop a type theory with a universe (\cref{subsec:SC+U}) of small types, $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).

@@

We extend the type theory in \cref{subsec:SC+U} with $\Pi$-types, i.e.\ dependent function types, whose the context of its codomain is extended with the domain type.
For the QIIT of type theory in \cite{Altenkirch2016a}, it amounts to adding the following constructors for the type formation rule, the term introduction/elimination rules:
\begin{alignat*}{3}
  \Pi     &: \implicit{\Gamma}           \; && (A : \Ty\,\Gamma) \to \Ty\,(\Gamma, A) \to \Ty\,\Gamma \\
  \lambda &: \implicit{\Gamma, A, B}     \; && \Tm\;(\Gamma, A)\;B \to \Tm\,\Gamma\,(\Pi\;A\;B) \\
  \mathsf{app} &: \implicit{\Gamma, A, B}\; && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B
\end{alignat*}
the computation rules
\begin{alignat*}{5}
  \Pi\beta           & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t) && =^{\Tm\;\Gamma\;A} && t \\
  \Pi\eta            & : \implicit{\Gamma, A, B, t}\; && t                          && =^{\Tm\;\Gamma\;A} && \lambda(\mathsf{app}\,t)
\end{alignat*}
and the structural rules for substitution
\begin{alignat}{5}
  []\Pi              & : \implicit{\Gamma, \Delta, \sigma, A, B}\;    && [ \sigma ]_{\Ty}\,(\Pi\;A\;B) && =^{\Ty\;\Gamma} && \Pi\,(\sub{\sigma}{A})\,([\sigma\uparrow A]\,B) \label{eq:def-type-subst-8} \\
  \mathsf{[]\lambda} & : \implicit{\Gamma, \Delta, \sigma, A, B, t}\; && [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma}_{\alert{[]\Pi}} && \lambda\,([\sigma \uparrow A]\,t) \nonumber
\end{alignat}
where $\sigma \uparrow A$ is the \emph{lifting}  of $\sigma$ by a type $A$ defined as $(\pi_1\idS; \sigma, \alert{\transfib{\Tm\;\Gamma}{[;]_T}{\color{black}\pi_2\idS}})$.

@@

Each constructor $c$ should be interpreted by a corresponding \emph{method} $c^M$ indexed by it, e.g.,

@@

The elimination principle says that the functions $\elim_{\Ctx}$, $\elim_{\Sub}$, $\elim_{\Ty}$ and $\elim_{\Tm}$ as soon as motives and methods are given. When the motive families are constant, we call the principle a \emph{recursion principle} instead.
We refer to \cref{appendix:full-def} and our Agda development for the full definition of the elimination principle.

@@

For instance, in \cref{ex:pi-comp-qiirt}, the identity $\pi_2(\sigma ; \tau) = \pi_2(\sigma; (\pi_1 \tau, \pi_2 \tau)$ could be proved by $\mathit{apd}_f \colon (p : x =^A y) \to f\,x =^{P}_{p} f\,y$ and the equality constructor $\pi\eta : \tau = (\pi_1\tau, \pi_2\tau)$, resulting an additional transport to remove later.
However, as $[\pi_1(\sigma; \tau)]_{\Ty}\,A$ and $[\pi_1(\sigma; (\pi_1\tau, \pi_2\tau))]_{\Ty}\,A$ are strictly equal, the resulting identity is indeed homogeneous.
Hence, we use the heterogeneous congruence $\mathit{hap}_f\colon (p : x \simeq y) \to f\,x\simeq f\,y$ (by \lstinline|hcong|) and switch back to the ordinary identity type $f\,x = f\, y$ (by \lstinline|≅-to-≡|) as their types match.

@@

In the future, we hope to settle the foundational status of the
combination of quotient types, inductive-inductive types, and the kind
of generalised inductive-recursive types we have considered here,
including overlapping patterns. In fact, our original motivation for
embarking on this work was seeking examples that could give evidence
that such a combination could be useful.
%
Various schemas for inductive types have been developed in recent
years~\cite{Ghani2017,Kaposi2018,Kovacs2023}, and it seems
promising to see if they can be extended to also handle quotient
inductive-inductive-recursive types. Overlapping patterns are
discussed by Cockx~\cite{Cockx2014}, but we would like to integrate it
with such a schema. Eremondi and Kammar~\cite{Eremondi2025} has
recently given a promising possible direction for overlapping patterns
using sheaf theory.
\LT{Perhaps also mention that the notion of models might be interesting, but it is out of the scope of this paper.}

>>>>

\subparagraph*{Plan of the paper}
In the rest of the paper, we introduce the metalanguage in \cref{sec:meta}.
Then, we develop type theories formulated as QIIRTs in \cref{sec:QIIRTs}, followed by their elimination principles and usage in \cref{sec:models}.
We outline the \Agda formal implementation in \cref{sec:formal-implementation}, and we conclude with related and future work in \cref{sec:conclusion}.

@@

We work in intensional Martin-L\"of type theory with uniqueness of identity proofs. In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality and $\eta$-rules for $\Pi$-types and $\Sigma$-types.
%
We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using \Agda notation $(x : A) \to B\,x$.
%
We are further inspired by \Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an `implicit' argument which should be clear from the context.
We write $\Set_i$ for the hierarchy of universes with the level index $i$ suppressed.

@@

An ordinary inductive type $A$, such as the type of natural numbers $A = \mathbb{N}$, allows functions out of the type $f : A \to X$ to be defined by recursion, using an elimination rule which says that such a function $f$ is defined by its values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of~$A$. For example, a function $f: \mathbb{N} \to X$ is given by the values $f(0)$ and $f(\suc\,n)$, where the latter might refer to the already computed value $f(n)$.
%
The type is first defined inductively by listing its constructors, and then afterwards recursive functions can be defined out of it.

In contrast, an inductive-recursive type is defined inductively \emph{simultaneously} with a recursively defined function out of it. The prototypical example is Martin-L\"of's universe \`a la Tarski, where a type $U$ of codes for small types is defined simultaneously with a `decoding' function $T : U \to \Set$ which maps each code to the type it represents. Crucially, the code for dependent function types
\[
  \pi : (a : U) \to (T(a) \to U) \to U
\]
with $T(\pi(a, b)) = (x : T\,a) \to T(b\, x)$ only makes sense because we already have access to the decoding function $T$.
%
Inductive-recursive types (IRT) were introduced by Dybjer~\cite{Dybjer2000}, and then extensively studied by Dybjer and Setzer~\cite{Dybjer1999,Dybjer2003}.
The general schema allows one to define a type $A$ inductively, together with a recursive function $F : A \to D$, for some fixed (possibly large) type $D$.

An IRT $(A, F)$ comes with an elimination rule that allows to define a dependent function  $f : (x : A) \to P(x)$ by giving values $f(\mathsf{c}_i\,x)$ on all of the constructors $\mathsf{c}_i$ of $A$, again reflecting that $A$ is inductively defined.
%
However, note that the recursive function $F : A \to D$ is \emph{not} defined using this elimination rule, but part of the definition of the type itself.
%
In particular, the elimination rule can be used to define endofunctions on $A$, but the recursive function  $F : A \to D$ must always target a fixed pre-existing type $D$.
%
In this work, we will consider an extension \emph{generalised} IRTs, where also $F : A \to A$ is allowed.
%
Such IRTs were previously used by Danielsson~\cite{Danielsson2006} to represent type theory in type theory, and are discussed by Kaposi~\cite{Kaposi2023}.
%
Their general metatheory would be interesting future work.

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020}, where constructors can introduce not only new elements but also new equalities between the introduced elements.
%
Since we assume the uniqueness of identity proofs, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type of \emph{join lists} $\JList{A}$ can be given by the QIT
with the following constructors:
\begin{alignat*}{3}
  \Jnil                    & : \implicit{A}      && \JList{A} \\
  \Jsing{\blank}           & : \implicit{A}      && A \to \JList{A} \\
  \Jconcat{\blank}{\blank} & : \implicit{A}      && \JList{A} \to \JList{A} \to \JList{A} \\
  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
  \mathsf{assoc}           & : \implicit{\ell_0, \ell_1, \ell_2}\; && \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
\end{alignat*}

A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a type family $B : A \to \Set$, where both $A$ and $B$ are defined inductively --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively.
%; indeed, inductive-inductive definitions only make sense when $D = \Set$.
The metatheory of QIITs has been extensively studied by Altenkirch, Kaposi and Kov\'acs~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.

We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: the type $\JList{A}$ is defined simultaneously with a recursion $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list. We keep the constructors $\Jnil$, $\Jsing{\blank}$ and $\Jconcat{\blank}{\blank}$, but replace $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
\[
  \mathsf{nf} : \implicit{A}\;(\ell, \ell' : \JList{A}) \to \flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell' \to \ell =^{\JList{A}} \ell'
\]
which states that join lists are equal as soon as their flattenings are equal. For $\flatten$ to be well defined, it needs to respect the $\mathsf{nf}$ constructor, but this is a tautology: if $\ell =^{\JList{A}} \ell'$ because $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$, then obviously $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$.

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
\begin{figure}[b]
\begin{minipage}[t]{.48\textwidth}
  \begin{alignat*}{5}
    \blank  & + \blank     && : \N \to \N \to \N \\
    0       &  + m         && = m  \\
  n         &  + 0         && = m  \\
  (\suc\;n) &  + m         && = \suc\;(n + m) \\
  n         &  + (\suc\;m) && = \suc\;(n + m)
\end{alignat*}
\end{minipage}
\begin{minipage}[t]{.5\textwidth}
\begin{alignat*}{5}
  \blank         & +'\blank     && : \N \to \N \to \N \\
  0              & +' m         && \defeq m  \\
  (\suc\;n)      & +' 0         && \defeq \suc\;n \\
  (\suc\;n)      & +' (\suc\;m) && \defeq \suc\;(n +' \suc\;m)
\end{alignat*}
\end{minipage}
\caption{Two seemingly different definitions of addition}
\label{fig:additions}
\end{figure}
Functions are typically defined informally using pattern matching for convenience and readability. Formally, however, they are either defined using an explicit pattern-matching construct or elaborated into case trees, which correspond to the \emph{eliminator}, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination and completeness checking.
This elaboration offers the conventional notation, while making computational behaviour opaque and favouring asymmetric definitions.
For example, two definitions of addition in \cref{fig:additions} are the same
after elaboration in \Agda: eliminating the first argument of $+$ distinguishes $0$ and $\suc\;n$ and refines the second and the third clauses into clauses of $+'$, making the fourth clause useless.
Each clause of $+'$ is strict (hence the symbol $\defeq$ is used) while clauses of $+$ are not necessarily strict by definition.
%This semantics for function definitions decides how definitions reduce and therefore how identities are proved.

Cockx~\cite{Cockx2014} explores an alternative semantics for pattern matching, where all clauses are strict, even in the presence of \emph{overlapping patterns}.
Thus, when a function $f$ is defined with overlapping patterns, the notation $f\;\overrightarrow{p_i} \reduce t_i$ is used to emphasize its reduction behaviour.
Under this semantics, each clause of $+$ is strictly equal to its right-hand side, also allowing the symmetry addition to be proved commutative directly by induction on $n$ and $m$.

@@

We begin with a recall of the definition of substitution calculus~\cite{Martin-Lof1992} as a QIIT, emphasising how the use of transports in its very definition complicates formal reasoning.
We then introduce its QIIRT counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning, and develop a type theory with a universe (\cref{subsec:SC+U}) of small types, $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).

@@

We extend the type theory in \cref{subsec:SC+U} with $\Pi$-types, i.e.\ dependent function types, where the context of its codomain is extended with the domain type.
For the QIIT of type theory in \cite{Altenkirch2016a}, it amounts to adding the following constructors for the type formation rule, the term introduction/elimination rules:
\begin{alignat*}{3}
  \Pi     &: \implicit{\Gamma}           \; && (A : \Ty\,\Gamma) \to \Ty\,(\Gamma, A) \to \Ty\,\Gamma \\
  \lambda &: \implicit{\Gamma, A, B}     \; && \Tm\;(\Gamma, A)\;B \to \Tm\,\Gamma\,(\Pi\;A\;B) \\
  \mathsf{app} &: \implicit{\Gamma, A, B}\; && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B
\end{alignat*}
the computation rules
\begin{alignat*}{5}
  \Pi\beta           & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t) && =^{\Tm\;\Gamma\;A} && t \\
  \Pi\eta            & : \implicit{\Gamma, A, B, t}\; && t                          && =^{\Tm\;\Gamma\;A} && \lambda(\mathsf{app}\,t)
\end{alignat*}
and the structural rules for substitution
\begin{alignat}{5}
  []\Pi              & : \implicit{\Gamma, \Delta, \sigma, A, B}\;    && [ \sigma ]_{\Ty}\,(\Pi\;A\;B) && =^{\Ty\;\Gamma} && \Pi\,(\sub{\sigma}{A})\,([\sigma\uparrow A]\,B) \label{eq:def-type-subst-8} \\
  \mathsf{[]\lambda} & : \implicit{\Gamma, \Delta, \sigma, A, B, t}\; && [\sigma]_{\Tm} (\lambda\,t) && =^{\Tm\,\Gamma}_{\alert{[]\Pi}} && \lambda\,([\sigma \uparrow A]\,t) \nonumber
\end{alignat}
where $\sigma \uparrow A$ is the \emph{lifting}  of $\sigma$ by a type $A$ defined as $(\pi_1\idS; \sigma, \alert{\transfib{\Tm\;\Gamma}{[;]_T}{\color{black}\pi_2\idS}})$.

@@

Each constructor $\mathsf{c}$ should be interpreted by a corresponding \emph{method} $\mathsf{c}^M$ indexed by it, e.g.,

@@

The elimination principle says that the functions $\elim_{\Ctx}$, $\elim_{\Sub}$, $\elim_{\Ty}$ and $\elim_{\Tm}$ as soon as motives and methods are given.
\LT[noinline]{says what?}
When the motive families are constant, we call the principle a \emph{recursion principle} instead.
We refer to \cref{appendix:full-def} and our \Agda development for the full definition of the elimination principle.

@@

For instance, in \cref{ex:pi-comp-qiirt}, the identity $\pi_2(\sigma ; \tau) = \pi_2(\sigma; (\pi_1 \tau, \pi_2 \tau)$ could be proved by $\apdfunc f \colon (p : x =^A y) \to f\,x =^{P}_{p} f\,y$ and the equality constructor $\pi\eta : \tau = (\pi_1\tau, \pi_2\tau)$, resulting an additional transport to remove later.
However, as $[\pi_1(\sigma; \tau)]_{\Ty}\,A$ and $[\pi_1(\sigma; (\pi_1\tau, \pi_2\tau))]_{\Ty}\,A$ are strictly equal, the resulting identity is indeed homogeneous.
Hence, we use the heterogeneous congruence $\mathit{hap}_f\colon (p : x \simeq y) \to f\,x\simeq f\,y$ (by \lstinline|hcong|) and switch back to the ordinary identity type $f\,x = f\, y$ (by \lstinline|≅-to-≡|) as their types match.

@@

In the future, we hope to settle the foundational status of the
combination of quotient types, inductive-inductive types, and the kind
of generalised inductive-recursive types we have considered here,
including overlapping patterns. In fact, our original motivation for
embarking on this work was seeking examples that could give evidence
that such a combination could be useful.
%
Various schemas for inductive types have been developed in recent
years~\cite{Ghani2017,Kaposi2018,Kovacs2023}, and it seems
promising to see if they can be extended to also handle quotient
inductive-inductive-recursive types. Overlapping patterns are
discussed by Cockx~\cite{Cockx2014}, but we would like to integrate it
with such a schema. Eremondi and Kammar~\cite{Eremondi2025} has
recently given a promising possible direction for overlapping patterns
using sheaf theory.
%\LT{Perhaps also mention that the notion of models might be interesting, but it is out of the scope of this paper.}
